Patterns and anti-patterns

### 2021-03-29 analysis/environments.Rmd
Analysis

---
title: "Environments and eval in R"
author: "Pierre Donat-Bouillud"
output:
  html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data
  calls_packages_path: /package/summarized.fst
---

@@

source("inc/latextags.R")
source("inc/setup.R")
source("inc/paths.R")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "environments.tex"), prefix = "", default = TRUE)

source("inc/insights.R", local = knitr::knit_global())
```

@@

```{r loading_fst}
eval_calls_path <- paste0(params$base_dir, params$calls_packages_path)

eval_calls <- read_fst(eval_calls_path) %>%
  as_tibble()

stopifnot(!is.na(eval_calls$eval_call_srcref))
```

@@

```{r packages}
nb_eval_calls <- count(eval_calls, wt = nb_ev_calls)

nb_call_sites <- nb_eval_call_sites(eval_calls)
```

There are `r count(eval_calls, wt=nb_ev_calls)$n` eval calls and `r eval_calls %>% select(eval_call_srcref) %>% n_distinct()` eval call sites in the dataset.

@@

```{r envir_default}
eval_calls_default <- eval_calls %>% filter(is.na(envir_expression))

nb_default <- count(eval_calls_default, wt = nb_ev_calls)
nb_default_sites <- nb_eval_call_sites(eval_calls_default)

eval_calls_explicit <- eval_calls %>% filter(!is.na(envir_expression))
nb_explicit <- count(eval_calls_explicit, wt = nb_ev_calls)
nb_explicit_sites <- nb_eval_call_sites(eval_calls_explicit)
```

@@

@@

nb_rcpp <- count(eval_calls_rcpp, wt = nb_ev_calls)
nb_rcpp_sites <- nb_eval_call_sites(eval_calls_rcpp)

@@

nb_env_list <- count(eval_calls_env_list, wt = nb_ev_calls)
nb_env_list_sites <- nb_eval_call_sites(eval_calls_env_list)
```

@@

nb_arg_caller <- count(eval_calls_arg_caller, wt = nb_ev_calls)
nb_arg_caller_sites <- nb_eval_call_sites(eval_calls_arg_caller)
```

@@

```{r}
eval_calls_arg_caller %>%
  count(envir_from_arg, wt = nb_ev_calls, sort = TRUE) %>%
  knitr::kable()
```

@@

nb_absolute <- count(eval_calls_absolute, wt = nb_ev_calls)
nb_absolute_sites <- nb_eval_call_sites(eval_calls_absolute)
```

@@

```{r}
eval_calls_absolute %>%
  filter(envir_distinct == 2) %>%
  select(eval_call_expression, environment_class, envir_distinct)
```

@@

```{r}
env_calls %>% filter(!is.na(new_env), specific_env == "global") %>% count(new_env, wt = nb_ev_calls, sort = TRUE)

env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(eval_source, wt = nb_ev_calls, sort = TRUE)
```

@@

nb_integer <- count(eval_calls_integer, wt = nb_ev_calls)
nb_integer_sites <- nb_eval_call_sites(eval_calls_integer)
```

@@

nb_null <- count(eval_calls_null, wt = nb_ev_calls)
nb_null_sites <- nb_eval_call_sites(eval_calls_null)
```

>>>>

---
title: "Environments and eval in R"
author: "Pierre Donat-Bouillud"
output:
  html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data/package/
  calls_path: summarized.fst
  corpus_path: corpus.txt
  static_path: evals-static.csv
  force_rebuild: FALSE
---

@@

source("inc/latextags.R")
source("inc/functions.R")
source("inc/paths.R")
theme_set(theme_minimal())

dataset_name <- basename(params$base_dir)

# Tags will be prefixed by the dataset name
create_tags(path(TAGS_DIR, paste0(dataset_name, "_environments.tex")), prefix = dataset_name, default = TRUE)
# Also use ggsave_prefix instead of ggsave if you need to save a graph

calls_path <- paste0(params$base_dir, params$calls_path)
corpus_path <- paste0(params$base_dir, params$corpus_path)
static_path <- paste0(params$base_dir, params$static_path)

lock_path <- paste0(params$base_dir, ".environments-lock.fst")
min_path <- paste0(params$base_dir, "Eenvir.fst")

nb_eval_call_sites <- function(dataset) {
  dataset %>% pull(src_ref) %>% n_distinct()
}

```

@@

```{r loading_fst}
tibble(package = read_lines(corpus_path)) -> C_raw
read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P

rebuild <- TRUE

E_envir <- NULL

if (file.exists(lock_path)) {
  saved <- read.fst(lock_path)[[1,1]]
  if (saved == file.size(calls_path)) {
    read_fst(min_path) %>% as_tibble() -> E_envir
    rebuild <- FALSE
  }
}

if(params$force_rebuild || rebuild) {
  read_fst(calls_path) %>% as_tibble() -> E_raw

  E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL" & is.na(envir_expression), "NULL", envir_expression))

  E_raw %>% select(
    run_package = package, #      The package which was run to trigger the eval
    ev_variant = eval_function, # Which eval variant was called
    duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
    src_ref = eval_call_srcref, # Source ref for the eval call site
    file, #                       Script file name (name of this run)
    ev_package = eval_source, #   Package of the call site
    type = expr_resolved_type_tag, #   Argument type
    ev_expr = expr_expression,
    n_op = interp_eval,
    envir = envir_expression,
    envir_type,
    caller_expression,
    environment_class
  ) -> E_envir

  E_envir %>% write_fst(min_path)
  sz <- file.size(calls_path) %>% as_tibble()
  write.fst(sz, lock_path)
}

C_raw -> C
corpus_size <- nrow(C)
```

@@

```{r packages}
nb_eval_calls <- count(E_envir, wt = duplicates)

nb_call_sites <- E_envir %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_eval_calls` eval calls and `r nb_call_sites` eval call sites in the dataset.

@@

```{r envir_default}
eval_calls_default <- E_envir %>% filter(is.na(envir))

nb_default <- count(eval_calls_default, wt = duplicates)
nb_default_sites <- nb_eval_call_sites(eval_calls_default)

eval_calls_explicit <- E_envir %>% filter(!is.na(envir))
nb_explicit <- count(eval_calls_explicit, wt = duplicates)
nb_explicit_sites <- nb_eval_call_sites(eval_calls_explicit)

overview_table(
  r("default envir sites", nb_default_sites),
  r("defaut envir sites percent", ratio(nb_default_sites, nb_call_sites)),
  r("non default envir sites", nb_explicit),
  r("non defaut envir sites percent", ratio(nb_explicit, nb_call_sites))
)
```

@@

# Most common envir arguments

We show the most frequent environment expression per site. There is only one environment expression per site, as the envir expression is captured before being resolved.

```{r envir_most_common}
frequent_envir <- E_envir %>% group_by(src_ref) %>% 
  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "))

stopifnot(frequent_envir$n_envir <= 1)
```

@@

nb_rcpp <- count(eval_calls_rcpp, wt = duplicates)
nb_rcpp_sites <- nb_eval_call_sites(eval_calls_rcpp)

@@

nb_env_list <- count(eval_calls_env_list, wt = duplicates)
nb_env_list_sites <- nb_eval_call_sites(eval_calls_env_list)
```

@@

nb_arg_caller <- count(eval_calls_arg_caller, wt = duplicates)
nb_arg_caller_sites <- nb_eval_call_sites(eval_calls_arg_caller)
```

@@

```{r}
eval_calls_arg_caller %>%
  count(envir_from_arg, wt = duplicates, sort = TRUE) %>%
  knitr::kable()
```

@@

nb_absolute <- count(eval_calls_absolute, wt = duplicates)
nb_absolute_sites <- nb_eval_call_sites(eval_calls_absolute)
```

@@

```{r}
eval_calls_absolute %>%
  filter(envir_distinct == 2) %>%
  select(ev_expr, environment_class, envir_distinct)
```

@@

```{r}
env_calls %>% filter(!is.na(new_env), specific_env == "global") %>% count(new_env, wt = duplicates, sort = TRUE)

env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(eval_source, wt = duplicates, sort = TRUE)
```

@@

nb_integer <- count(eval_calls_integer, wt = duplicates)
nb_integer_sites <- nb_eval_call_sites(eval_calls_integer)
```

@@

nb_null <- count(eval_calls_null, wt = duplicates)
nb_null_sites <- nb_eval_call_sites(eval_calls_null)
```

### 2021-03-29 analysis/patterns2.Rmd
Analysis

@@

```{r}
parse_eval <- E_cano %>% filter(type == "expressions" | str_detect(ev_expr, fixed("parse")))

@@

As a first approximation, it corresponds to type `language`.

```{r}
parse_eval <- E_cano %>% filter(type == "language")

## selecting columns in a dataframe

@@

# Non-necessary `eval`

@@

>>>>

The diversity of non-default environments is higher than for other patterns.

@@

```{r}
#parse_eval <- E_cano %>% filter(type == "expressions" | str_detect(ev_expr, fixed("parse")))
parse_eval <- E_cano %>% filter(!is.na(parsed))

@@

As a first approximation, it corresponds to type `language` and the usage of `substitute`. Indeed, using `quote` or `bquote` to get a language expression is the responsability of the caller of function, not of the function itself. On the contrary, with `substitute` and laziness, it is possible to capture the AST of an argument.

```{r}
language_eval <- E_cano %>% filter(type == "language")

nb_language_eval <- language_eval %>% pull(src_ref) %>% n_distinct()

There are `r nb_language_eval` sites with language expressions.

We can already detect if substitute is present directly in the `eval` site:

```{r}
substitute_eval <- language_eval %>% filter(str_detect(ev_expr, fixed("substitute")))
nb_substitute_eval_sites <- substitute_eval %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_substitute_eval_sites` sites that have `eval` inside their non-resolved expression.

To also detect the case when `substitute` is used before the call to `eval`, we can approximate it by looking whether a function with `eval` also contains a call to substitute before it.

## Selecting columns in a dataframe

@@

#  `eval` anti-patterns

@@

See normalized expression notebook.
